---
title: >-
  Context specific functional scores for protein-protein interaction networks
author:
- name: Ana Galhoz
  affiliation: hmgu
  email: ana.galhoz@helmholtz-muenchen.de
  correspondance: yes
- name: Denes Turei
  affiliation: unihd
institute:
- hmgu: Helmholtz Zentrum Muenchen
- unihd: Institute for Computational Biomedicine, Heidelberg University
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
package: wppi
abstract: |
  The `wppi` package calculates context specific scores for genes in
  the network neighborhood of genes of interest. The context specificity
  is ensured by the selection of the genes of interest and potentially by
  using a more relevant subset of the ontology annotations, e.g. selecting
  only the diabetes related categories. The PPI network and the functional
  annotations are obtained automatically from public databases, though
  it's possible to use custom databases. The network is limited to a
  neighborhood of the genes of interest. The ontology annotations are also
  filtered to the genes in this subnetwork. Then the adjacency matrix is
  weighted according to the number of common neighbors and the similarity
  in functional annotations of each pair of interacting proteins. On this
  weighted adjacency matrix a random walk with returns is performed. The
  final score for the genes in the neighborhood is the sum of their scores
  (probabilities to be visited) in the random walk.
  The method can be fine tuned by setting the neighborhood range, the
  restart probability of the random walk and the threshold for the random
  walk.
vignette: |
  %\VignetteIndexEntry{WPPI workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
fig_width: 9
fig_height: 7
---

```{r suppress-progressbars, echo=FALSE}
options(progress_enabled = FALSE, width = 80)
```

# Dependencies

The `wppi` package depends on the `OmnipathR` package. Since it relies on
features more recent than the latest Bioconductor version (OmnipathR 2.0.0
in Bioconductor 3.12), until the release of Bioconductor 3.13, it is
recommended to install OmnipathR from git:

```{omnipath, results='hide', eval=FALSE}
require(devtools)
install_github('saezlab/OmnipathR')
library(OmnipathR)
```

# Complete workflow in a single call

The `score_candidate_genes_from_PPI` function executes the full workflow.
The only mandatory input is a set of genes of interest.

```{r workflow, results='hide'}
library(wppi)
# example gene set
genes_interest <- c(
    'ERCC8', 'AKT3', 'NOL3', 'TTK',
    'GFI1B', 'CDC25A', 'TPX2', 'SHE'
)
scores <- score_candidate_genes_from_PPI(genes_interest)
scores
# # A tibble: 295 x 3
#    score gene_symbol uniprot
#    <dbl> <chr>       <chr>
#  1 0.247 KNL1        Q8NG31
#  2 0.247 HTRA2       O43464
#  3 0.247 KAT6A       Q92794
#  4 0.247 BABAM1      Q9NWV8
#  5 0.247 SKI         P12755
#  6 0.247 FOXA2       Q9Y261
#  7 0.247 CLK2        P49760
#  8 0.247 HNRNPA1     P09651
#  9 0.247 HK1         P19367
# 10 0.180 SH3RF1      Q7Z6J0
# # . with 285 more rows
```

# Workflow step by step

## Database knowledge

The database knowledge is provided by `wppi_data`. By default all directed
protein-protein interactions are used from OmniPath. By passing various
options the network can be customized. See more details in the documentation
of the `OmnipathR` package, especially the
`import_post_translational_interactions` function. For example, to use only
the literature curated interactions one can use the `datasets = 'omnipath'`
parameter:

```{r database-knowledge-omnipath}
omnipath_data <- wppi_omnipath_data(datasets = 'omnipath')
```

The `wppi_data` function retrieves all database data at once. Parameters to
customize the network can be passed directly to this function.

```{r database-knowledge, results='hide'}
db <- wppi_data(datasets = c('omnipath', 'kinaseextra'))
names(db)
# [1] "hpo"      "go"       "omnipath" "uniprot"
````

Optionally, the Human Phenotype Ontology (HPO) annotations relevant in the
context can be selected. For example, to select the annotations related to
diabetes:

```{r database-knowledge-hpo}
# example HPO annotations set
HPO_data <- wppi_hpo_data()
HPO_interest <- unique(dplyr::filter(HPO_data, grepl('Diabetes', Name))$Name)
```

## Converting the interactions to an igraph graph object

To work further with the interactions we first convert it to an `igraph`
graph object:

```{r omnipath-graph}
graph_op <- graph_from_op(db$omnipath)
```

## Subgraph from the neighborhood of genes of interest

Then we select a subgraph around the genes of interest. The size of the
subgraph is determined by the range of this neighborhood (`sub_level`
argument for the `subgraph_op` function).

```{r neighborhood-subnetwork, results='hide'}
graph_op_1 <- subgraph_op(graph_op, genes_interest)
igraph::vcount(graph_op_1)
# [1] 256
```

## Weighted adjacency matrix

The next step is to assign weights to each interaction. The weights are
calculated based on the number of common neighbors and the similarities of
the annotations of the interacting partners.

```{r weighted-adjacency-matrix}
w_adj <- weighted_adj(graph_op_1, db$go, db$hpo)
```

## Random walk

The random walk with restarts algorithm uses the edge weights to score the
overall connections between pairs of genes. The result takes into accound
also the indirect connections, integrating the information in the graph
topology.

```{r random-walk}
w_rw <- random_walk(w_adj)
```

## Scoring proteins

At the end we can summarize the scores for each protein, taking the sum of
all adjacent connections. The resulted table provides us a list of proteins
prioritized by their predicted importance in the context of interest
(disease or condition).

```{r scoring-proteins}
scores <- prioritization_genes(graph_op_1, w_rw, genes_interest)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```